\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2020-10-27 manuscript.Rnw
Some updates to iCTSTM theory section

\subsection{Cohort discrete time state transition models} \label{sec:cDTSTMs}
cDTSTMs simulate the probability that a cohort of patients is in each of $H$ health states over time. Time is measured at discrete times with each time point known as a model cycle. A $1 \times H$ state vector that stores the probability of being in each health state at time $t$ is written as $v_t = (v_{1t}, v_{2t}, \ldots, v_{Ht})$ where $\sum_i v_{ht} = 1$. A transition probability matrix is denoted by $P_t$ where the $(r,s)$th element represents a transition from state $r$ to state $s$ between times $t$ and $t+1$. The state vector at time $t+1$ for each of $T$ model cycles can be computed as,

\begin{equation} \label{eqn:markov-sim}
v_{t+1}= v_t P_t, \quad t = 0,\ldots, T.
\end{equation}

@@

\subsection{Individual continuous time state transition models} \label{sec:iCTSTMs}
iCTSTMs simulate individual trajectories between health states in continuous time using random number generation. Trajectories are simulated for multiple patients and costs and QALYs are computed by averaging across the simulated patients. A reasonably large number of patients must be simulated to ensure that expected values are stable \citep{ohagan2007monte}.

Simulation is performed using a parametric and flexible parametric multi-state modeling framework based on the \pkg{flexsurv} package \citep{jackson2016flexsurv}. Let $t_j$ denote the time a patient enters state $r$ and $t_{j+1}$ denote the time a patient starting in state $r$ transitions to state $s$. Time-to-event $t^*$ for the $r \rightarrow s$ has probability density function,

@@

where parameters $\theta = (\theta_1, \theta_2, \ldots, \theta_p)$ may depend on covariates $z_p$ through the link function $g(\theta_p) = z_p^T \gamma$ where $\gamma$ are a vector of regression coefficients. In a time inhomogeneous Markov model, time $t^*=t_{j+1}$ is conditional on not experiencing event $s$ until time $t_j$ (i.e., it is left-truncated at time $t_j$). In a semi-Markov model, time-to-event is expressed in terms of $t^* = t_{j+1}-t_j$ and a patient enters state $s$ at time $t_j + t^*$.

@@

\begin{algorithm}
\caption{Simulation of individual continuous time state transition model}
\label{alg:sim-ictstm}
\begin{enumerate}
\item Let $r$ be the starting state and starting time be $t_j = 0$. The number of permitted transitions from state $r$ is given by $n_r$.
\item Simulate times $\mathcal{T} = \{t_{1, j+1}, t_{2, j+1}, \ldots t_{n_r, j+1}\}$ to each of the $n_r$ permitted transitions.
\item Set the time of the transition $t_{j+1}$ equal to the minimum simulated time in $\mathcal{T}$ and the next state $s$ to the state with the minimum simulated time.
\item Set $r=s$ and $t_j = t_{j+1}$. If the patient is still alive, repeat the previous steps until death.
\end{enumerate}
\end{algorithm}

>>>>

\subsection{Cohort discrete time state transition models} \label{sec:cDTSTMs}
cDTSTMs simulate the probability that a cohort of patients is in each of $H$ health states over time. Time is measured at discrete times with each time point known as a model cycle. A $1 \times H$ state vector that stores the probability of being in each health state at time $t$ is written as $x_t = (x_{1t}, x_{2t}, \ldots, x_{Ht})$ where $\sum_i x_{ht} = 1$. A transition probability matrix is denoted by $P_t$ where the $(r,s)$th element represents a transition from state $r$ to state $s$ between times $t$ and $t+1$. The state vector at time $t+1$ for each of $T$ model cycles is given by,

\begin{equation} \label{eqn:markov-sim}
x_{t+1}^T= x_t^T P_t, \quad t = 0,\ldots, T.
\end{equation}

@@

\subsection{Individual continuous time state transition models} \label{sec:iCTSTMs}
iCTSTMs simulate individual trajectories between health states using random number generation. Trajectories are simulated for multiple patients and costs and QALYs are computed by averaging across the simulated patients. A reasonably large number of patients must be simulated to ensure that expected values are stable \citep{ohagan2007monte}.

In continuous time, a patient is in state $X(t)$ at time $t$. State transitions are modeled using a multi-state modeling framework \citep{putter2007tutorial} where the probability of a transition from state $r$ to state $s$ is governed by the hazard function,

\begin{equation}
h_{rs}(t) = \lim_{\Delta t\to 0} \frac{\Prob(X(t + \Delta t) = s | X(t) = r)}{\Delta t}.
\end{equation}

Simulated disease progression is characterized by $J$ distinct jumps between health states $D = \{(t_0, X(t_0)), (t_1, X(t_1)), \ldots (t_J, X(t_J))\}$ with a patient remaining in a health state from time $t_j$ until transitioning to the next state at $t_{j+1}$. Jumps between health states are simulated using parametric and flexible parametric survival models as implemented in the \pkg{flexsurv} package \citep{jackson2016flexsurv}. Specifically, if a patient enters state $r$ at time $t_j$, then a probability density function for the time-to-event $t^*$ for the $r \rightarrow s$ transition is,

@@

where parameters $\theta = (\theta_1, \theta_2, \ldots, \theta_p)$ may depend on covariates $z_p$ through the link function $g(\theta_p) = z_p^T \gamma$ and $\gamma$ is a vector of regression coefficients. In a time inhomogeneous Markov model, time $t^*=t_{j+1}$ is conditional on not experiencing event $s$ until time $t_j$ (i.e., it is left-truncated at time $t_j$). In a semi-Markov model, time-to-event is expressed in terms of $t^* = t_{j+1}-t_j$ and a patient enters state $s$ at time $t_j + t^*$.

@@

\begin{algorithm}
\caption{Simulation of individual continuous time state transition model}
\label{alg:sim-ictstm}
\begin{enumerate}
\item Let $r$ be the state entered at time $t_j$. The number of permitted transitions from state $r$ is given by $n_r$. If $j=0$, then $t_j = 0$. 
\item Simulate times $\mathcal{T} = \{t_{1, j+1}, t_{2, j+1}, \ldots t_{n_r, j+1}\}$ to each of the $n_r$ permitted transitions.
\item Set the time of the transition $t_{j+1}$ equal to the minimum simulated time in $\mathcal{T}$ and the next state $s$ to the state with the minimum simulated time.
\item Set $r=s$ and $t_j = t_{j+1}$. If the patient is still alive, repeat the previous steps until death.
\end{enumerate}
\end{algorithm}
